---
title: "Child Care"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
runtime: shiny
resource_files:
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.cpg
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.dbf
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.prj
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbn
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbx
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shp.xml
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(extrafont)
library(readxl)
library(janitor)
library(lubridate)

loadfonts()
```

```{r loading jon docs and shapefiles, cache=TRUE, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", size = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", size = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial"))

regions <- read_csv("Data/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("Data/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         edr = ifelse(Name == "Minnesota", "Minnesota", edr) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         mif = ifelse(is.na(mif), "TC", mif),
         mif = as.factor(mif),
         mif = fct_relevel(mif, "NW", "NE", "WC", "EC", "SW", "SE", "TC"))

color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("Data/Shapefiles/county shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)

```

```{r master child care, include=FALSE, cache=TRUE}
master.child.care <- read_csv("Data/Child care/Master-child-care.csv") %>%
  mutate(Dem_Desc = fct_relevel(Dem_Desc, "Entirely rural", "Town/rural mix", "Urban/town/rural mix", "Entirely urban"),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))
  

```

<br>

# Child care capacity{.tabset}

The charts below provide the change in child care capacity (by type of care) from 2019 to 2020. 

The first thing you will notice is that family-based care took a beating during COVID with significant declines across the board. Interestingly, the largest declines in family-based care took place in our more urban counties.

Breaking the data down by RUCA county groups shows that our entirely rural counties had the largest decline in child care capacity between 2019 and 2020. Although these counties experienced declines in both center care and family care capacity, the largest decline was in center care. All other county groups experienced increases in center based care which helped stave off larger declines in total child care capacity.

The hardest hit region was Northeast Minnesota, where they were also the only region to experience a decline in center based care capacity. All other regions experienced declines in total care capacity but were minimized by increases in center based care.

Breaking the data down by EDR shows some interesting nuance. In the northern regions of Minnesota, EDR 2 - Headwaters was able to minimize declines in total child care capacity with an increase in center-based care of 22.6%. EDR 4 - West Central was able to have an increase in child care capacity with a 5.3% increase in center-based care as well. A similar story occurred in EDR 6E - Southwest Central where they were able to have an overal capacity increase of 1.3% thanks to an increase in center based care of 7.5%. Lastly, EDR 6W - Upper MN Valley saw a massive decline in center based care of 24.3%.

Long story short - family care declined throughout Minnesota while in some places an increase in center based care was able to stave off larger total child care declines.


<br>

```{r prep child care capacity, include=FALSE, cache=TRUE}

cc.capacity.county <- master.child.care %>%
  filter(year %in% c(2019, 2020)) %>%
  mutate(tot.capacity = fcc.capacity + ccc.capacity) %>%
  gather(key = "care.type", value = "capacity", 9:11) %>%
  mutate(care.type = str_replace(care.type, "ccc.capacity", "Center care capacity"),
         care.type = str_replace(care.type, "fcc.capacity", "Family care capacity"),
         care.type = str_replace(care.type, "tot.capacity", "Total care capacity")) 

cc.capacity.county.map <- cc.capacity.county %>%
  group_by(county, care.type) %>%
  mutate(capacity.annual.change = (capacity - lag(capacity)) / lag(capacity)) %>%
  ungroup() %>%
  filter(year != 2019,
         care.type == "Total care capacity") %>%
  mutate(capacity.annual.change.bins = cut(capacity.annual.change,
                                            breaks = c(-1, -.1, -.05, 0, .5, 1),
                                            labels = c("-10% or more", "-10% to -5%", "-5% to 0%", "0% to 5%", "More than 5%"))) %>%
  left_join(mn_counties[,c(4,7)], by = "countyfp")


cc.capacity.ruca <- cc.capacity.county %>%
  group_by(Dem_Desc, care.type, year) %>%
  summarise(capacity = sum(capacity)) %>%
  ungroup() %>%
  group_by(Dem_Desc, care.type) %>%
  mutate(capacity.index = (capacity - capacity[year == min(year)]) / capacity[year == min(year)],
         capacity.annual.change = (capacity - lag(capacity)) / lag(capacity)) %>%
  ungroup() %>%
  mutate(data_id = as.character(seq(n())),
         year = as_factor(year)) %>%
  filter(year != "2019")
  
cc.capacity.pr <- cc.capacity.county %>%
  group_by(planning.region, care.type, year) %>%
  summarise(capacity = sum(capacity)) %>%
  ungroup() %>%
  group_by(planning.region, care.type) %>%
  mutate(capacity.index = (capacity - capacity[year == min(year)]) / capacity[year == min(year)],
         capacity.annual.change = (capacity - lag(capacity)) / lag(capacity)) %>%
  ungroup() %>%
  mutate(data_id = seq(n()),
         year = as_factor(year)) %>%
  filter(year != "2019")

cc.capacity.edr <- cc.capacity.county %>%
  group_by(edr, planning.region, care.type, year) %>%
  summarise(capacity = sum(capacity)) %>%
  ungroup() %>%
  group_by(edr, planning.region, care.type) %>%
  mutate(capacity.index = (capacity - capacity[year == min(year)]) / capacity[year == min(year)],
         capacity.annual.change = (capacity - lag(capacity)) / lag(capacity)) %>%
  ungroup() %>%
  mutate(data_id = seq(n()),
         year = as_factor(year)) %>%
  filter(year != "2019")
  
```

## RUCA

<br>

```{r chart child care capacity index ruca}
cc.capacity.ruca.plot <- ggplot(cc.capacity.ruca, aes(Dem_Desc, capacity.annual.change, fill = Dem_Desc)) +
  facet_wrap(~care.type, ncol = 2) +
  geom_hline(yintercept = 0, color = "black") +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(Dem_Desc, "\nYear: ", year, "Child care type: ", care.type, "\nCare capacity: ", comma(capacity, accuracy = 1), "\nChange in capacity since pervious year: ", percent(capacity.annual.change, accuracy = .1), sep = ""))) +
  labs(x="", y = "", color="", title = "Change in child care capacity 2019 - 2020", caption = "MN Department of Health")+
  geom_label(data = cc.capacity.ruca, position = position_dodge(width = .9), aes(label = percent(capacity.annual.change, accuracy = .1)), show.legend = FALSE, size = 3) +
  scale_y_continuous(labels=scales::percent) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
  theme_line+
  scale_fill_manual(values = color.ruca,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 18))

  girafe(ggobj = cc.capacity.ruca.plot, width_svg = 10, height_svg = 10) %>%
    girafe_options(opts_selection(type = "none"),
                   opts_toolbar(saveaspng = FALSE),
                   opts_sizing(rescale = FALSE))

```

<br>

## Planning Region

<br>

```{r chart child care capacity index pr}
cc.capacity.pr.plot <- ggplot(cc.capacity.pr, aes(planning.region, capacity.annual.change, fill = planning.region)) +
  facet_wrap(~care.type, ncol = 2) +
  geom_hline(yintercept = 0, color = "black") +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(planning.region, "\nYear: ", year, "Child care type: ", care.type, "\nCare capacity: ", comma(capacity, accuracy = 1), "\nChange in capacity since pervious year: ", percent(capacity.annual.change, accuracy = .1), sep = ""))) +
  labs(x="", y = "", color="", title = "Change in child care capacity 2019 - 2020", caption = "MN Department of Health")+
  geom_label(data = cc.capacity.pr, position = position_dodge(width = .9), aes(label = percent(capacity.annual.change, accuracy = .1)), show.legend = FALSE, size = 3) +
  scale_y_continuous(labels=scales::percent) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
  theme_line+
  scale_fill_manual(values = color.pr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 18))

  girafe(ggobj = cc.capacity.pr.plot, width_svg = 10, height_svg = 10) %>%
    girafe_options(opts_selection(type = "none"),
                   opts_toolbar(saveaspng = FALSE),
                   opts_sizing(rescale = FALSE))

```

<br>

## EDR{.tabset}

<br>

### North

<br>

```{r chart child care capacity index edr}
cc.capacity.edr.n.plot <- ggplot(filter(cc.capacity.edr, planning.region %in% c("Northwest", "Northeast")), aes(edr, capacity.annual.change, fill = edr)) +
  facet_wrap(~care.type, ncol = 2) +
  geom_hline(yintercept = 0, color = "black") +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(edr, "\nYear: ", year, "Child care type: ", care.type, "\nCare capacity: ", comma(capacity, accuracy = 1), "\nChange in capacity since pervious year: ", percent(capacity.annual.change, accuracy = .1), sep = ""))) +
  labs(x="", y = "", color="", title = "Change in child care capacity 2019 - 2020", caption = "MN Department of Health")+
  geom_label(data = filter(cc.capacity.edr, planning.region %in% c("Northwest", "Northeast")), position = position_dodge(width = .9), aes(label = percent(capacity.annual.change, accuracy = .1)), show.legend = FALSE, size = 3) +
  scale_y_continuous(labels=scales::percent) +
  scale_x_discrete(guide = guide_axis(n.dodge = 3)) +
  theme_line+
  scale_fill_manual(values = color.edr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 18))

  girafe(ggobj = cc.capacity.edr.n.plot, width_svg = 10, height_svg = 10) %>%
    girafe_options(opts_selection(type = "none"),
                   opts_toolbar(saveaspng = FALSE),
                   opts_sizing(rescale = FALSE))

```

<br>

### Middle

<br>

```{r chart child care capacity index edr middle}
cc.capacity.edr.m.plot <- ggplot(filter(cc.capacity.edr, planning.region %in% c("Central", "Seven County Mpls-St Paul")), aes(edr, capacity.annual.change, fill = edr)) +
  facet_wrap(~care.type, ncol = 2) +
  geom_hline(yintercept = 0, color = "black") +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(edr, "\nYear: ", year, "Child care type: ", care.type, "\nCare capacity: ", comma(capacity, accuracy = 1), "\nChange in capacity since pervious year: ", percent(capacity.annual.change, accuracy = .1), sep = ""))) +
  labs(x="", y = "", color="", title = "Change in child care capacity 2019 - 2020", caption = "MN Department of Health")+
  geom_label(data = filter(cc.capacity.edr, planning.region %in% c("Central", "Seven County Mpls-St Paul")), position = position_dodge(width = .9), aes(label = percent(capacity.annual.change, accuracy = .1)), show.legend = FALSE, size = 3) +
  scale_y_continuous(labels=scales::percent) +
  scale_x_discrete(guide = guide_axis(n.dodge = 3)) +
  theme_line+
  scale_fill_manual(values = color.edr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 18))

  girafe(ggobj = cc.capacity.edr.m.plot, width_svg = 10, height_svg = 10) %>%
    girafe_options(opts_selection(type = "none"),
                   opts_toolbar(saveaspng = FALSE),
                   opts_sizing(rescale = FALSE))

```

<br>

### South


<br>

```{r chart child care capacity index edr south}
cc.capacity.edr.s.plot <- ggplot(filter(cc.capacity.edr, planning.region %in% c("Southwest", "Southeast")), aes(edr, capacity.annual.change, fill = edr)) +
  facet_wrap(~care.type, ncol = 2) +
  geom_hline(yintercept = 0, color = "black") +
  geom_col_interactive(position = "dodge", aes(data_id = data_id, tooltip = paste(edr, "\nYear: ", year, "Child care type: ", care.type, "\nCare capacity: ", comma(capacity, accuracy = 1), "\nChange in capacity since pervious year: ", percent(capacity.annual.change, accuracy = .1), sep = ""))) +
  labs(x="", y = "", color="", title = "Change in child care capacity 2019 - 2020", caption = "MN Department of Health")+
  geom_label(data = filter(cc.capacity.edr, planning.region %in% c("Southwest", "Southeast")), position = position_dodge(width = .9), aes(label = percent(capacity.annual.change, accuracy = .1)), show.legend = FALSE, size = 3) +
  scale_y_continuous(labels=scales::percent) +
  scale_x_discrete(guide = guide_axis(n.dodge = 3)) +
  theme_line+
  scale_fill_manual(values = color.edr,
                     guide = guide_legend(ncol = 3)) +
  theme(legend.position = "bottom",
        text = element_text(size = 18))

  girafe(ggobj = cc.capacity.edr.s.plot, width_svg = 10, height_svg = 10) %>%
    girafe_options(opts_selection(type = "none"),
                   opts_toolbar(saveaspng = FALSE),
                   opts_sizing(rescale = FALSE))

```

<br>

## County map

<br>

```{r chart capacity change county map}
cc.capacity.county.map.plot <- ggplot(cc.capacity.county.map) +
  geom_sf_interactive(color = "grey85", aes(geometry = geometry, fill = capacity.annual.change.bins, data_id = countyfp, tooltip = paste(county, "\nTotal capacity in 2020: ", comma(capacity, accuracy = 1), "\nChange in capacity from 2019: ", percent(capacity.annual.change, accuracy = .1), sep = ""))) +
  theme_sf +
  scale_fill_manual(values = brewer.pal(n = 6, "RdYlBu")) +
  labs(title = "Change in child care capacity 2019 - 2020", caption = "\nMN Department of Health") +
  theme(legend.box.margin = margin(50, 0, 0, -100),
        text = element_text(size = 18))

girafe(ggobj = cc.capacity.county.map.plot, height_svg = 10, width_svg = 10) %>%
  girafe_options(opts_sizing(rescale = FALSE))


```

<br>

