---
title: "Summary"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
runtime: shiny
resource_files:
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.cpg
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.dbf
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.prj
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbn
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.sbx
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shp.xml
- Data/Shapefiles/County shapefiles/MNCounties_MNDOT.shx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(sf)
library(ggrepel)
library(scales)
library(shiny)
library(shinycssloaders)
library(ggiraph)
library(kableExtra)
library(rmapshaper)
library(cowplot)
library(DT)
library(htmlwidgets)
library(RColorBrewer)
library(extrafont)
library(readxl)
library(janitor)
library(lubridate)

loadfonts()
```

```{r loading jon docs and shapefiles, cache=TRUE, include=FALSE}
theme_bar <- theme_bw() +
  theme(panel.grid.major = element_line(color = "grey70", size = 0.1),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.key.size = unit(1, "lines"),
        legend.margin = margin(0,0,0,0),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        text = element_text(family = "Arial"))

theme_line <- theme_bw() +
  theme(legend.background = element_rect(fill = "transparent", color = "transparent"),
        legend.key = element_rect(fill = "transparent"),
        legend.text = element_text(margin = margin(l = 2)),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "grey70", size = 0.1),
        axis.ticks = element_blank(),
        axis.text = element_text(face = "bold"),
        panel.border = element_blank(),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial"))


theme_sf <- theme_bw() +
  theme(axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "white"),
        panel.border = element_blank(),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_blank(),
        legend.text = element_text(margin = margin(l = 2)),
        legend.margin = margin(0,0,0,0),
        legend.key.size = unit(1, "lines"),
        text = element_text(family = "Arial"))

regions <- read_csv("Data/Join docs/county_regions.csv") %>%
    select(5,6) %>%
    unique() %>%
    mutate(edr = str_replace(edr, "  ", " "),
           planning.region = str_replace(planning.region, " Minnesota", ""),
           planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
           edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"))

counties.regions <- read_csv("Data/Join docs/county_regions.csv") %>%
  rename(mif = `MIF Region`) %>%
  mutate(countyfp = formatC(countyfp, width = 3, flag = "0"),
         Name = str_to_title(Name),
         Name = str_replace(Name, "Q", "q"),
         Name = str_replace(Name, "Of The", "of the"),
         Name = str_replace(Name, "Mcleod", "McLeod"),
         Dem_Desc = ifelse(Name == "Minnesota", "Minnesota", Dem_Desc) ,
         edr = str_replace(edr, "  ", " "),
         planning.region = str_replace(planning.region, " Minnesota", ""),
         planning.region = fct_relevel(planning.region, "Northwest", "Northeast", "Central", "Seven County Mpls-St Paul", "Southwest", "Southeast"),
         edr = fct_relevel(edr, "EDR 1 - Northwest", "EDR 2 - Headwaters", "EDR 3 - Arrowhead", "EDR 4 - West Central", "EDR 5 - North Central", "EDR 6E- Southwest Central", "EDR 6W- Upper Minnesota Valley", "EDR 7E- East Central", "EDR 7W- Central", "EDR 8 - Southwest", "EDR 9 - South Central", "EDR 10 - Southeast", "EDR 11 - 7 County Twin Cities", "Minnesota"),
         mif = ifelse(is.na(mif), "TC", mif),
         mif = as.factor(mif),
         mif = fct_relevel(mif, "NW", "NE", "WC", "EC", "SW", "SE", "TC"))

color.ruca <- c("Entirely rural" = "#009933", "Town/rural mix" = "#99CC33", "Urban/town/rural mix" = "#CC9966", "Entirely urban" = "#754C29", "Minnesota" = "black")

color.pr <- c("Northwest" = 	"#4575b4", "Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black")

color.edr <- c("EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365", "Minnesota" = "black")

color.pr.edr <- c ("Northwest" = "#4575b4","Northeast" = "#e0f3f8", "Central" = "#fee090", "Seven County Mpls-St Paul" = "#d73027", "Southwest" = "#91bfdb", "Southeast" = "#fc8d59", "Minnesota" = "black", "EDR 1 - Northwest" = "#b3cde3", "EDR 2 - Headwaters" = "#8c96c6", "EDR 3 - Arrowhead" = "#fe9929", "EDR 4 - West Central" = "#8856a7", "EDR 5 - North Central" = "#810f7c", "EDR 6E- Southwest Central" = "#e5f5f9", "EDR 6W- Upper Minnesota Valley" = "#bdc9e1", "EDR 7E- East Central" = "#99d8c9", "EDR 7W- Central" = "#2ca25f", "EDR 8 - Southwest" = "#74a9cf", "EDR 9 - South Central" = "#0570b0", "EDR 10 - Southeast" = "#d7301f", "EDR 11 - 7 County Twin Cities" = "#d8b365")

mn_counties <- st_read("Data/Shapefiles/county shapefiles/MNCounties_MNDOT.shp", quiet = TRUE) %>%
  ms_simplify(keep = .01, keep_shapes = TRUE) %>%
  rename(countyfp = FIPS_CODE)

```

<br>

Below is a summary of each section. You can navigate to the main sections using the tabs above.

<br>

# Infections 

**Infection rate was highest in our more city/rural areas of the state**

Interestingly, the cumulative cases per 100,000 shows that the counties with the highest COVID rates were in our urban/town/rural county group (22,424 per 100,000) and our town/rural county group (21,998 per 100,000). These were following by entirely urban county group (20,324 per 100,000) and our entirely urban group (18,568 per 100,000). One would think that the counties with the densest population would have the highest case rate. However, one theory is that both the urban/town/rural counties and the town/rural counties were just dense enough to spread the virus while also being more reluctant to take precautions to stop the spread. To test this idea, the seven county metro was removed from the "Entirely urban" county group and sure enough, the entirely urban county group (sans 7-county metro) had the highest cases per 100,000. 

**High infection rates in greater Minnesota could be do to industry make-up**

Infections per 100,000 in greater Minnesota were highest in counties that have a large manufacturing employment sector (including food processing). Counties such as Kandiyohi, Roseau, Nobles, and counties surrounding Waseca and Mower had the highest cases per 100,000 in the entire state. However, these counties also have a high employment in manufacturing which were kept open during the beginning of the pandemic due to their "essential" status. Large spreader events occurred in these plants which likely drove up the case count.


**Highest death rates from COVID are in our rural areas**

Most interesting is that although our most rural areas had the lowest cumulative cases per 100,000, they unfortunately had the highest death rate of all counties. Entirely rural counties had a death rate of 275 per 100,000 throughout the pandemic. The lowest was in our entirely urban counties with 188 deaths per 100,000. This is most likely due to the fact that rural areas have a larger percentage of their population that is older, sicker, and have a higher percentage of people with co-morbidity. Interestingly, the same counties that had the highest case rates (Kandiyohi, Roseau, etc...) did not have the highest death rates indicating that the virus spread through a healthier population in those counties. In addition to these factors, the higher death rate in our rural areas likely has to do with the fact mentioned in our next highlight....

**Significantly lower vaccination rates in rural areas**

The highest vaccination rate exists in our entirely urban counties with around 75% of their population having the full series of shots. Our rural areas had between 60% and 65% of their population with the full series of shots. The lowest percentages exist in the counties just north of the twin cities metro where many of those counties barely have 50% of their population vaccinated with not even one shot. There were some rural regions that were surprising however. Counties in Southwest and Northeast in particular had some of the highest vaccination rates in Greater Minnesota.

**The lowest vaccination rates are among American Indian populations**

The percentage of American Indians vaccinated with at least one dose was the lowest among all race and ethnicity with 60.3% and only 53.5% with the complete vaccination series. Asian and Pacific Islanders have the highest with over 85% with at least one dose followed by whites and Hispanic with 70.5%.

<br>

# Health care

**Hospitals in our urban and regional centers bared the greatest burden in treating COVID patients**

The percentage of beds occupied by COVID patients in rural areas throughout the pandemic were surprisingly low. But this is likely due to coordination among hospitals and within health care systems to transport COVID patients to the larger regional hospitals. We will have to talk to Santo to get a better handle on this.

**Rural hospitals were in a precarious financial state immediately before the pandemic**

Although we don't yet have 2020 financial data, the 2019 hospital annual reports showed that hospitals in our rural areas had average profit losses compared to more metropolitan hospitals. 

<br>

# Economic impacts

**Job vacancies are still crazy high, maybe even worse than pre-pandemic**

Although there was a slight drop in the job vacancy rate during the pandemic, it continued to still be high across rural Minnesota. And, there was an enormous jump in the job vacancy rate during the first half of 2021 as companies ramped up hiring again. During 2020 the job vacancy rate hovered between 4% and 5.5% (3% is considered healthy), the first half of 2021 saw job vacancy rates increase to between 8% and 10%.

**Drop in people participating in the labor force hit rural regions hardest**

One of the larger negative impacts from the pandemic is the number of people that have left the labor force. Although the drop in the number of people participating in the labor force occurred in all of Minnesota, rural areas were the hardest hit. Rural regions which were already experience little to no growth in their labor force, such as EDRs in the Southwest and Northwest, now have less people participating in the labor force today than they did in 2012!

**The drop in employment worst among females workers in rural areas**

Although employment (note - this isn't labor force) decreased among all genders, the largest decreases occurred among female workers, but only outside of the urban areas. For example, in 2021 the Northwest region had 1% more males employed compared to 2012 yet -3.1% less females employed. And this drop occurred right in 2020 as the previous years saw increases in employment among females. The opposite is true in urban areas where there was a larger drop in male employment compared to female employment.

**This is not a typical economic downturn**

Although the number of people participating in the labor force declines along with overall employment, wages continue to go up significantly. This is not typical of previous economic downturns where employment and wages would decline together. What this tells us is that people are choosing to not work (for many reasons) even though there are plenty of jobs. To try and entice people into the labor pool, employers are increasing wages. Interestingly, the largest decreases in employment have occurred in rural Minnesota while these same regions have experienced the largest wage growth.

**The new remote work fad not occurring as much outside of the seven county metro**

Much has been made about the rise of telecommuting for employment. A report released by MN DEED analyzed job openings and found that there was a rather large disparity in the percentage of jobs allowing telecommuting by region. The seven-county metro had by far the largest percentage of job openings offering telecommuting (13%) followed by Northeast (8.2%) and Southeast (7.5%). It then drops considerably to the remaining regions with Northwest at (2.4%), Central (2.0%), and Southwest (2.0%).

**Significant difference in local options sales tax from Minneapolis**

A sampling of cities that are collecting local option sales tax revenues shows little to no negative impact from the pandemic and in fact may have been helpful. All cities showed rather normal revenue trends in 2021 compared to 2020 while also seeing larger than expected INCREASES in revenues as the pandemic wore on through May and June of 2021. The opposite is true for Minneapolis where LOST fell off a cliff.

<br>

# Child care

**COVID impacted child care capacity more in urban areas**

Surprisingly, our most rural areas of the state experienced a 
